Efficiency

### 2018-09-14 Second.Rnw
a couple more bits

The competence of RADM Sammut's program management and the technical advisors on whom he depends is further marred by the comment about lithium-ion batteries being now firmly excluded from the design because the ``RAN does not want to be introducing a massive number of new technologies simultaneously".  This is telling because lead-acid batteries are known to have a relatively high internal resistance, which increases as they become hot, which occurs during levels of faster discharge.  This means that the efficiency of the batteries falls substantially at higher rates of discharge, reducing substantially the total amount of stored energy that they can usefully recover at very high discharge rates.  Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms efficiency) at higher rates of discharge when  compared to lead-acid batteries.  It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.  Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.  Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.  This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to under one hour.  With lithium ion batteries, it is plausible that slightly over 2 hours at this speed could be sustained.

As I have also pointed out

@@

>>>>

The competence of RADM Sammut's program management and the technical advisors on whom he depends is further marred by the comment about lithium-ion batteries being now firmly excluded from the design because the ``RAN does not want to be introducing a massive number of new technologies simultaneously".  This is telling because lead-acid batteries are known to have a relatively high internal resistance, which increases as they become hot, which occurs during levels of faster discharge.  This means that the efficiency of the batteries falls substantially at higher rates of discharge, reducing substantially the total amount of stored energy that they can usefully recover at very high discharge rates.  Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms efficiency) at higher rates of discharge when  compared to lead-acid batteries.  It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.  Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines, at least if the case is to be made that priorisation of that extremely high-speed sprint capability should be made above all other performance characteristics, which I have argued is an obscure and implausible case to be made \ref{SENmorrison}  Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.  This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to well under one hour.  With lithium ion batteries, it is plausible that around 2 hours at this speed might be sustained.  The exclusion of lithium batteries \textit{because} of the pumpjet's inclusion further indicates that RADM Sammut remains completely unaware of such interactions between technologies.

@@

\section{The weight and size argument}
Both Dr David Kershaw, Chief of Maritime Division at DSTG\parencite{kerr2018}, as well as Mr Brent Clark\parencite{clark2018} have both made arugments to the effect of suggesting that it was the Australian submarine's very large size which is what made the pumpjet viable, where it might not be for smaller submarines.  I responded directly to Clark's comments arguing that the size necessitated by the extra fuel and batteries a pumpjet demanded could hardly be counted as an advantage \parencite{morrison2018b}.  However, Dr Kershaw's argument reveals a different as to why it was that size was important.  His comments make it clear that it is actually the large weight of the pumpjet \textit{itself} rather than the additional fuel and batteries it required, that was significant.

This argument implies that a smaller submarine couldn't adequately provide sufficient buoyancy or balance to support the heavy pumpjet. If the submarine couldn't provide such buoyancy or balance, the consequences would be catastophic failure, specifically the submarine would tend to either tail-stand, or sink.  It would be unthinkable for a competent manufacturer, conscious of the weight of the pumpjet and such risks, to neglect to account for such risks.  Weights and balances are known and predictable quantities, which could most easily be accounted for in the design of a new submarine.

However, Dr Kershaw's argument is most comprehensively debunked by the known fact that pumpjets have indeed been successfully retro-fitted to existing smaller diesel-electric submarines.  As one would expect, none of the catastrophic failures eventuated.  It is reported that the Russian Kilo Class submarine B-871 'Alrosa' has and does take part in exercises, and is remains in active service.  However, given that no other conventional submarines have since been designed with or converted to pumpjet propulsion, it is difficult to conclude that the benefits of a pumpjet on this type of submarine outweigh the costs.  It is trivial, however, to conclude that the weight of the pumpjet itself is not prohibitive for a smaller submarine, including  one smaller than the Collins Class, such as the Kilo.

Furthermore, the French also fitted a pumpjet to an Agosta 70 class submarine for a trial, as confirmed by Brent Clark in the Senate \ref{SENmorrison}. This submarine is even smaller than the Kilo, around 1500 tonnes, exactly the type of 'coastal' submarine that Brent Clark said a pumpjet wouldn't make sense for \parencite{clark2018}.  One would assume however that if the weight of the pumpjet itself was likely to cause the submarine to tail-stand or sink, such a trial would not have even been attempted.

Further physical arguments could be made as to how the weight of the pumpjet ought to scale with the size of submarine.  There are obvious reasons why the width/height (which must be equal, since the jet is cylindrical/conical) ought to scale in proportion to the diameter of the hull, in order to engage a proportionally larger water-column to produce the thrust necessary.  Similarly, the length  of the shround is likely to remain relatively fixed in proportion to the diameter for a given performance requirement. It is unlikely that the proportional thickness of the blades or shroud could be greatly diminished at larger scales without structural integrity and hydrodynamic qualities of the shroud and rotor blades being compromised.  All of these relationships strongly suggest that Dr Kershaw's implied suggestion that the weight of the pumpjet does not scale consistently with the weight of the submarine is completely unfounded, probably thoughtless, or otherwise  deliberately misleading.  However, given the clear-cut evidence that the type of failure that an over-weight propulsor would cause have been easily avoided on \textit{existing} smaller conventional submarines, such arguments wont be expanded in detail here.  It is abundantly clear that the arguments are entirely baseless, and demonstrably so.

It is difficult to imagine what sort of circumstance would lead both an acting CEO of the contracted design partner, who was aware of the experience of the French experiment with the Agosta, as well as a very senior defence scientist to make such demonstrably false claims in public.  In the case of Dr Kershaw, it can be seen that he deliberately and repeatedly (at the conclusion of his interview  \parencite{kerr2018}) drew attention to the size being significant.  His concluding remarks for the interview were that ``The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role...". To deliberately mount an argument that is so simply ridiculed and dismissed as being so significant seems to be a bizarre choice for a supposedly competent scientist to make.  If stronger arguments were available, one would assume that they would be made instead. With over a million dollars already spent on modelling and experiments, one would assume that if the results of that research could confidently refute the conclusions my own research \parencite{morrison2018}, such a desperate and embarrassing statement would not need to be made.

In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independed of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, technical, yet understandable truth which is used to justify a completely blatant technical falsehood.  Even if Dr Kershaw is not familiar with this area of science and speaking outside his competence, it is difficult to see how such a pattern of argument occuring so consistently by accident.  For Dr Kershaw to genuinely believe in the arguments that he is making here is almost unthinkable.  For him to casually or accidentally make such careless arguments also seems implausibly reckless.  A more likely explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desparate attempt to deflect attention in order to avoid admitting as much.

\section{The interaction between the hull and the propulsor}
The argument that the 'hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss critcism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}. It has also been referred to in an answer to a Question on Notice given by Senator Ketter\ref{QON1}.  The answer to the Question on Notice, and also Dr Kershaw's comments comprise clearly the more detailed explanations, which  will be referred to in principal in this section.   However, the reference of RADM Sammut to this line of argument on multiple occasions, almost a year separated both before and after the publication of my paper\parencite{morrison2018} and my senate evidence \ref{SENmorrison} clearly indicates that it is a significant, perhaps the most significant line of defence of the  pumpjet in the program chief's mind.  The written response supplied by Defence to the Senate Economic References Committee also refers clearly to this argument \parencite{defence2018}.

As has already been noted, there are glaring contradictions from a project-management perspective which oppose this argument.  In particular, given that we are, according to RADM Sammut accept the pumpjet that was offered by France, and our submarine supposedly has a near identical hull diameter and  length  to the nuclear Barracuda, it bizarre that any adjustment at all of either the aft section of the submarine, or the pumpjet itself, ought to be contemplated. The major advantage that the Naval Group proposal \textit{ought} to have is that those potentially complicated engineering challenges can be safely and entirely avoided.  Furthermore, the idea that the hull is what necessitates the tuning, as opposed to the speed profile, is almost incomprehensibly naive, as I pointed out in my evidence given to the Senate \ref{SENmorrison}.  It is noteworthy that no officials from Naval Group have ever alluded to this 'tuning' argument, or any adjustments to be made to the pumpjet or the hull on account of one another.  However, it seems pertinent here to give a reasonable discussion of the relevant physics nonetheless, in order that the public might fully assess the sort of argument being made by our most senior defence officials.

It is true that the hull does indeed change the flow of incoming water to a propulsor.  The area behind the vessel where the water is on being slightly dragged along with the ship (or the water is moving less quickly past the vessel, in the vessel's reference frame) is known as the wake.  Since any aft-mounted propulsor has to operate in the wake-field, it is true that wake is of importance to the design of the propolsor.  What is not true, however, is that this is grounds for disregarding open-water efficiency, or that it was neglected from consideration in the preparation of my report.

In reality, the operation of a propulsor in the wake-field is an absolutely standard consideration for deriving propulsive efficiencies, as can be seen in any number of standard texts \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}.

There is a ratio which is commonly discussed, relative rotative efficiency (denoted $\eta_R$), which has as it's precise purpose to describe the ratio between the efficiency of the propulsor in open water ($\eta_O$), and the efficiency of the propulsor in the wake of the hull ($\eta_B$).

It is widely acknowledged (for example in all of references \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}) that relative rotative efficiency generally lies around 1, and does not vary greatly.  For instance section 16.3.2, page 411 of reference \parencite{Mollard2011} says ``$\eta_R$ ranges typically from 0.95 to 1.05 and is often assumed to be unity for preliminary design purposes." \footnote{Observers may note that efficiencies are generally thought to range between 0 and 1.  The possibility that $\eta_R$ could be greater than one is opened up by the fact that it is actually a \textit{relative} efficiency, one efficiency divided by another.  Some propellers can actually work marginally better in the wake field than outside of it.}

In clear contrast however, the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.  In \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.  It is consequently vastly more important to give detailed consideration to early in the design and selection of a propulsor, particularly for a diesel-electric submarine.  Unless defence is proposing that a pumpjet can have a relative rotative efficiency of more than ~\%200, there is no feasibly means that this line of argument can sustain their dismissal of my conclusion.

A particularly desparate line of argument might lead one to argue that such arguments do not apply to submarines propulsors due to their great level of depth and the unique shape of their hull. This would not be true.  Owing to their generally cylindical shape and generally conical afterbody, submarines tend to have wake fields are much less complex and intense in the degree of their variation that surface ships. This is because ships bodies also have wave-making effects involved as well as a hull that tapers above the propulosor, creating an uneven effect with different heights.  In contrast, wake fields of propellers have a relatively even gradient in wake fraction (the intensity of waterspeed difference due to the hull) in a radial direction, with only the control surfaces ahead of the propulsor causing substantial variations.  Furthermore, at very low speeds, where pumpjets were shown to have the lowest efficiecies, the effect of the wake field would be the least compared to the open-water case, since the size of the boundary layer about the aft of the hull would be substantially diminished.  On absolutely all accounts, the interaction between the hull of a submarine and propulsor at low speeds cannot be credibly argued to have any impact which would overturn the conclusion reached about pumpjets having clearly and substantially inferior efficiency at low speed, as argued in \parencite{morrison2018}.

As such, the reason that my earlier paper did not explicitly discuss relative rotative efficiency and the impact of the wake field, this was due to it being so obviously and demonstrably of no bearing whatsoever on my conclusions, and I had assume that as much would immediately clear to any technically competent reviewer.

The use of this argument displays a degree of ignorance of the physics, as clearly described in numerous and publically available sources, that it is difficult to account for by incompetence alone.  A substantial measure of contempt for public intelligence must be required in order to appeal to such trivially simple statements of truth in order to support conclusions which can be so trivially demonstrated to be completely false.

To use a simple analogy to, imagine that you came to buy a coffee from a your favourite cafe, only to find that they weren't selling any coffee that day, and wouldn't be selling any more indefinitely.  Their coffee maching had been removed, and they were only selling cold-drip, coffee, stored in little bottles in the fridge. Overcoming initial shock (since you weren't aware of any other cafes who would make such a bizarre decision as to not sell espresso-coffee as a substantial part of their offering), you made some tentative inquiries as to why.  At first, you're told vaguely that it was to do with health and safety.  This seemed a little odd, since there obvious benefits to everyone's enjoyment and presumeably the business, which you point out to the cafe owner, and you ask how the health and safety risk could really be so large as to outweigh that.  With just a little annoyance at having to explain, the cafe owner briefly says something about temperature being the problem.  You look at the empty spot on the bench where the coffee machine used to be, and speculate that perhaps there really was some serious fault with the machine, maybe even a flaw in its design requireing it to be recalled.  Perhaps it was running uncontrollably hot, and might have risked melting componentry, creating electrocution risk, or even a risk of explosion if the boiler rose to an elevated pressure.  That would explain the safety.  Or perhaps there was some emerging science about carcinogens being produced in the high-temperature extraction of espresso, which cold-drip might presumeably avoid?  In anycase, it sounded like it could be very technical, and you decided to trust your very experienced and seemingly intelligent barista and forget about your lingering doubts as to why no other cafes were following suit.  Then later you find a notice posted up at the front of the cafe, explaining fully the rationale behind the abandmonment of the coffee machine, and the connection with temperature. In it, it is explained that all espresso-based coffees were hot, those with milk around 65 or 70 degrees, and long blacks or fresh espresso potentially above 90 degrees.  If consumed quickly or accidentally spilled such drinks could potentially cause burns.  Therefore in the interest of health and safety no more hot drinks would be served, and cold-drip would be the only coffee available.

You read the statement again, not quite believing that a hazard so obviously known and successfully navigated by millions of coffee drinkers every day, could be the cause of such a decision.  Althought it's true the owner didn't lie about it being a health and safety concern, and can't dispute the facts about the coffee temperatures being above 65 degrees, you still conclude that owner is either insane, or trying to disguise some other reason for abandoning the espresso coffee.
